---
title: "Credit card defaults prediction"
author: "Andrea"
date: "21 June 2020"
output: pdf_document
toc: yes
header-includes:
- \usepackage{makeidx}
- \makeindex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

In this project, using a dataset containing six months information, we want to create a model that predicts credit card defaults. From a risk management perspective, an accurate prediction and assessment of the credit risk is of relevant importance for the bank issuing the credit card.
The dataset we are going to use can be downloaded at the following url: \newline
\newline
"https://archive.ics.uci.edu/ml/machine-learning-databases/00350/default%20of%20credit%20card%20clients.xls".\newline
\newline
We have downloaded and manually saved a version of the *.xlsx* file in the "data" subfolder you can find within the github repo link we have provided together with the three files or at the following ulr: \newline
\newline
"https://github.com/andrearoetti/Choose-Your-Own/tree/master/data". \newline
\newline
Before getting started, we load the necessary packages for our work, after installing them in case they are not installed yet.
\newline
```{r installing, message = FALSE, warning = FALSE}
if (!require(tidyverse)) install.packages('tidyverse')
if (!require(lattice)) install.packages('lattice')
if (!require(caret)) install.packages('caret')
if (!require(readxl)) install.packages('readxl')
library(tidyverse)
library(lattice)
library(caret)
library(readxl)
```

We therefore use here below a relative path in order to read the excel file in R: for the sake of simplicity, we will simply call the dataset **credit_card**.
\newline
```{r read the file, message = FALSE}
credit_card <- read_xlsx("data/default of credit card clients.xlsx")
```
# Data cleaning and exploratory analysis

We first take a look of the first six rows to get a flavour of how the dataset is made.
\newline
```{r head}
head(credit_card)
```

We immediately notice that 

